
# Wet-Dry Switches

Before downsampling the data let's calculate the number of switches between wet and dry states in a 10min interval.
This will be a binary indicator (wets>0 or wets==0) that we can use along with the proportions.

Our current DEG files look like this:
```
date,time,wets0-20
2017-09-20,00:08:08,0
2017-09-20,00:18:08,0
2017-09-20,00:28:08,0
2017-09-20,00:38:08,0
2017-09-20,00:48:08,0
2017-09-20,00:58:08,0
2017-09-20,01:08:08,0
```

To calculate switches between 10min intervals, we need to transform wet values to a binary 

Example //

If we have a vector of wets equal to:
```[0, 20, 18, 0, 4, 2]```

The binary would be:
```[0, 1, 1, 0, 1, 1]```
        
Then we need to calculate the difference between each value.
  
  ```diff(c(0, 1, 1, 0, 1, 1))```
  
  Output:
  ```[1, 0, -1, 1, 0]```

Finally, we don't want negatives, so we can count absolute values
  
  ```abs(c(1, 0, -1, 1, 0))```
  
  Final Output:
  ```[1, 0, 1, 1, 0]```

The function below does this for one file, adding a column to our DEG files that tracks switches

```{r}
calc_switch <- function(file_path) {
  deg_data <- read_csv(file_path, show_col_types=FALSE)
  bird_id <- str_extract(basename(file_path), "(?<=filtered_)[A-Z-0-9]{5}")
  
  #calc numb. of switches per file
  switches_data <- deg_data %>%
    mutate(
      #switches between wet (non-zero) and dry (zero) states
      #row_number() assigns a unique number to each row to reference row positions
      #if it is the first row, set state to 0 since there are no rows to compare with
      #for all other rows (else case), calculate the numb. of switches 
      switches = ifelse(
        row_number() == 1, 0, 
        abs(diff( #calculates absolute values for state changes (1 or -1 depending on direction)
          ifelse(wets > 0, 1, 0)))),  #if wets>0 use 1, if not use 0 (transforms wets to binary)
      
      bird_id = bird_id)

  return(switches_data)
}
```

Then we can apply that function to all our files in our `clipped` directory
```{r}

#define deg files to use
deg_files <- list.files(output.dir, pattern = "*.deg", full.names = TRUE)#[1:5] #add if you want to subset

plan(multisession, workers = 4)

switch_metrics <- future_map_dfr(deg_files, calc_switches)

plan(sequential)

```


