# Proportions Wet

    Within each 2hr period we need to assess:
    •	Proportion wet or time wet ((Count wets in 2hr period * 30sec) / 72000sec)


Calculate proportions for downsampled data where the sum of wets was calculated in 2hr bins

## Calclate Proportions
 
Function to calculate proportion of wet time on the downsampled data:
```{r}

calc_prop <- function(deg_data, bin_time = bin_time) {
  
  bird_id <- deg_data$bird_id[1]
  
  #start by calculating metrics for a single 2hr bin
  prop_data <- deg_data %>%
    mutate(
      bird_id = bird_id,
      #proportion of time wet in bin_time period (wets*30 converts wetness level into secs spent wet)
      # (600sec per 10min interval) / 20 = 30 seconds per unit 
      prop_wet = (wets * 30) / (bin_time * 3600) #convert wets to secs and divide by bin_length in secs
      )

  return(prop_data)
}
```


Applying the function to our DEG files downsampled by 2hrs using sum of wets:

```{r}
plan(multisession, workers = 4)

bin_time = 2

deg_props_2hrs <- future_map_dfr(deg_2hrs, calc_prop, bin_time = bin_time)

deg_props_2hrs

```

Add the `colony` and `deployment_period` values from `metrics_md` to `deg_props_2hrs`

```{r}
deg_props_2hrs <- deg_props_2hrs %>%
  left_join(metrics_md %>% select(bird_id, deployment_period, colony), by = "bird_id")
```


