# Visualizing Metrics
    
    From Katie's Notes:
    
    Within each 2hr period we need to assess:
    •	Proportion wet or time wet ((Count wets in 2hr period * 30sec) / 72000sec)
    •	Number of switches between 0 and non-zero value
    
    •	Total time when entire 10min sampling interval was dry (count 0s * 10min)
    •	Total time when entire 10min sampling interval was wet (count >=1 * 10min)
    
    •	Longest continuous time dry (longest number of consecutive rows with wets==0  * 10min)
    •	Longest continuous time wet (longest number of consecutive rows with wets>=1 * 10min)


I start with setting up some plotting functions so we can look at different subsets of our data easily. Then i apply those plotting functions to the Baccalieu (2017-2018) DEG files (6 files) to visualize the metrics we discussed.

## Plotting Functions

### Plot Proportions

For proportions, I use the downsampled data by 2hr bins (taking the **sum** of wets).

```{r}
#plot by 2hr bins
plot_props <- function(data, y_var, y_label, plot_title) {
 plot_ly(data = data,
        x = ~as.POSIXct(paste(date, start_time), format = "%Y-%m-%d %H:%M:%S", tz = "UTC"),
        y= as.formula(paste0("~", y_var)), 
       type = 'scatter',
       mode = 'lines+markers',
       color = ~bird_id,
       line = list(width = 0.5),
       marker = list(size = 3)) %>%
        layout(
          title = list(text=plot_title, font=list(size=11)),
          xaxis = list(title = "Time (2-hour bins)", tickformat = "%b %Y", tickangle = 45),
          yaxis = list(title = y_label, rangemode = "tozero"), 
          legend = list(title = list(text = "Bird ID"), orientation = "v"),
          margin = list(t=100, b=100, l=100, r=100)) %>%
    config(responsive = TRUE)

  }

#or plot by day
plot_daily_props <- function(data, y_var, y_label, plot_title) {
 plot_ly(data = data,
        x = ~date,
        y= as.formula(paste0("~", y_var)), 
       type = 'scatter',
       mode = 'lines+markers',
       color = ~bird_id,
       line = list(width = 0.5),
       marker = list(size = 3)) %>%
        layout(
          title = list(text=plot_title, font=list(size=11)),
          xaxis = list(title = "Time (days)", tickformat = "%b %Y", tickangle = 45),
          yaxis = list(title = y_label, rangemode = "tozero"), 
          legend = list(title = list(text = "Bird ID"), orientation = "v"),
          margin = list(t=50, b=50, l=100, r=100))%>%
    config(responsive = TRUE)
  }
```


### Plot Switches

Plotting switches in **10min intervals** across a whole deployment period takes up a lot of mem and it's hard to visualise. Let's summarize the number of switches into a daily period to visualise, taking the sum of switches per day:

```{r}
daily_switches <- deg_switches %>%
  group_by(bird_id, date, deployment_period, colony) %>%
  summarize(daily_switches = sum(switches, na.rm = TRUE), 
            .groups = "drop")

#or plot by day
plot_daily_switches <- function(data, y_var, y_label, plot_title) {
 plot_ly(data = data,
        x = ~date,
        y= as.formula(paste0("~", y_var)), 
       type = 'scatter',
       mode = 'lines+markers',
       color = ~bird_id,
       line = list(width = 0.5),
       marker = list(size = 3)) %>%
        layout(
          title = list(text=plot_title, font=list(size=11)),
          xaxis = list(title = "Date", tickformat = "%b %Y", tickangle = 45),
          yaxis = list(title = y_label, rangemode = "tozero"), 
          legend = list(title = list(text = "Bird ID"), orientation = "v"),
          margin = list(t=100, b=100, l=100, r=100))%>%
    config(responsive = TRUE)
}

```


### Plot Longest Periods

These are the longest continuous dry and wet periods on a timeline (geom_segment)
```{r}

plot_longest_prds <- function(data, plot_title) {
  
  plot <- ggplot(data, aes(y=bird_id)) +
    geom_segment(aes(x = longest_dry_start, xend = longest_dry_end, yend = bird_id), 
                 size = 1, color = "lightblue") +  
    geom_segment(aes(x = longest_wet_start, xend = longest_wet_end, yend = bird_id), 
                 size = 1, color = "blue") +
    
    labs(title = plot_title,
         subtitle = "Light Blue: Dry | Blue: Wet",
         x = "Time", y = "Bird ID") +
    
    scale_x_datetime(breaks = scales::date_breaks("1 month"), labels = scales::date_format("%b %Y")) +
    
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 10),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  plotly_plot <- ggplotly(plot) %>%
    layout(margin = list(l = 100, r = 100, b = 100, t = 100),
           xaxis = list(tickangle = 45))%>%
    config(responsive = TRUE)
  
  return(plotly_plot)
}
  
```



### Plot Total Wets 

•	Total time when entire 10min sampling interval was dry (count 0s * 10min)
•	Total time when entire 10min sampling interval was wet (count >=1 * 10min)


```{r}
plot_total_times <- function(data, plot_title) {
  plot_ly(data = data) %>%
    add_trace(
      x = ~bird_id,
      y = ~total_time_wet,
      type = 'bar',
      name = 'Total time wet (mins)',
      marker = list(color = 'blue')
    ) %>%
    add_trace(
      x = ~bird_id,
      y = ~total_time_dry,
      type = 'bar',
      name = 'Total time dry (mins)',
      marker = list(color = 'lightblue')
    ) %>%
    layout(
      title = list(text = plot_title, font = list(size = 10)),
      xaxis = list(title = "Bird ID", tickangle = 45),
      yaxis = list(title = "Total time (mins)", rangemode = "tozero"),
      barmode = 'group',
      legend = list(title = list(text = "Metric")),
      margin = list(t = 100, b = 100, l = 100, r = 100)%>%
    config(responsive = TRUE)
    )
}
```


_____________________________________________________________________


Let's see what we have to work with per colony/deployment_periods
```{r}
deployment_summary
```



## Baccalieu - Ned Walsh 2017-2018

Let's start by visualizing switches/proportions for 2017-2018 files from Baccalieu (6 files total)

### Running plotting functions

**Proportions:**

```{r}
#subset the dataframe so we have less data to work with
bacc_2017_props <- deg_props_2hrs %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id, date, start_time, end_time, prop_wet)

head(bacc_2017_props)
```


```{r}
plot_bacc2017_props <- plot_props(data = bacc_2017_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (Baccalieu Island 2017-2018)")

```


This kind of hurts to look at so we can also try aggregating the data to days taking the sum of prop_wet.

**Daily Proportions:**

```{r}
bacc_2017_daily_props <- bacc_2017_props %>%
  group_by(bird_id, date) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")


plot_bacc2017_daily_props <- plot_daily_props(data = bacc_2017_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (Baccalieu Island 2017-2018)")

```


**Daily Switches:**

```{r}
#subset the dataframe so we have less data to work with
bacc_2017_switches <- daily_switches %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id, date, daily_switches)

head(bacc_2017_switches)
```

```{r}
plot_bacc2017_daily_switches <- plot_daily_switches(
  data = bacc_2017_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for Baccalieu Island (2017-2018)")

```


**Longest Wet/Dry Periods**

```{r}
bacc_2017_longest_prds <- deg_longest_prds %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)
```

```{r}
plot_bacc_2017_longest_prds <- plot_longest_prds(
  data = bacc_2017_longest_prds,
  plot_title = "Longest Wet and Dry Periods - Baccalieu Island 2017-2018")

```



**Total Wet/Dry**

```{r}

bacc_2017_total_wetdry <- deg_total_wetdry %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plot_bacc_2017_total_times <- plot_total_times(bacc_2017_total_wetdry, "Total Wet and Dry Times (Baccalieu Island, 2017-2018)")
```


### Final Plots 

```{r}
plot_bacc2017_props
```

```{r}
plot_bacc2017_daily_props
```

```{r}
plot_bacc2017_daily_switches
```

```{r}
plot_bacc_2017_total_times
```

```{r}
plot_bacc_2017_longest_prds
```