# Visualizing Metrics
    
    From Katie's Notes:
    
    Within each 2hr period we need to assess:
    •	Proportion wet or time wet ((Count wets in 2hr period * 30sec) / 72000sec)
    •	Number of switches between 0 and non-zero value
    
    •	Total time when entire 10min sampling interval was dry (count 0s * 10min)
    •	Total time when entire 10min sampling interval was wet (count >=1 * 10min)
    
    •	Longest continuous time dry (longest number of consecutive rows with wets==0  * 10min)
    •	Longest continuous time wet (longest number of consecutive rows with wets>=1 * 10min)


I start with setting up some plotting functions so we can look at different subsets of our data easily. Then I apply those plotting functions to various sets of the DEG files (6 files) to visualize the metrics we discussed (for Baccalieu, Kent's Island, and random files).

```{r include=FALSE}

source(here::here("project_settings.R"), echo=TRUE)

```

```{r}
#Load the plotting functions 
source(here::here("utils/plotting_functions.R"), echo=TRUE)
```
_____________________________________________________________________


Let's see what we have to work with per colony/deployment_periods
```{r}
deployment_summary
```



## Baccalieu Island 2017-2018

Let's start by visualizing switches/proportions for 2017-2018 files from Baccalieu (6 files total)

### Running plotting functions

**Proportions:**

```{r}
#subset the dataframe so we have less data to work with
bacc_2017_props <- deg_props_2hrs %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
    select(bird_id, date, start_time, end_time, prop_wet, colony, deployment_period)

head(bacc_2017_props)
```


```{r}
plotly_bacc2017_props <- plot_props(data = bacc_2017_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (Baccalieu Island 2017-2018)")

```


This kind of hurts to look at so we can also try aggregating the data to days taking the sum of prop_wet.

**Daily Proportions:**

```{r}
bacc_2017_daily_props <- bacc_2017_props %>%
  group_by(bird_id, date, colony, deployment_period) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")


plotly_bacc2017_daily_props <- plot_daily_props(data = bacc_2017_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (Baccalieu Island 2017-2018)")

```


**Daily Switches:**

```{r}
#subset the dataframe so we have less data to work with
bacc_2017_switches <- daily_switches %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id, colony, date, daily_switches, deployment_period)

plotly_bacc2017_daily_switches <- plot_daily_switches(
  data = bacc_2017_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for Baccalieu Island (2017-2018)")

```


**Longest Wet/Dry Periods**

```{r}
bacc_2017_longest_prds <- deg_longest_prds %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)


plotly_bacc_2017_longest_prds <- plot_longest_prds(
  data = bacc_2017_longest_prds,
  plot_title = "Longest Wet and Dry Periods - Baccalieu Island 2017-2018")

```



**Total Wet/Dry**

```{r}

bacc_2017_total_wetdry <- deg_total_wetdry %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plotly_bacc_2017_total_times <- plot_total_times(bacc_2017_total_wetdry, "Total Wet and Dry Times (Baccalieu Island, 2017-2018)")
```


### Final Plots 

```{r}
plotly_bacc2017_props
```

```{r}
plotly_bacc2017_daily_props
```

```{r}
plotly_bacc2017_daily_switches
```

```{r}
plotly_bacc_2017_total_times
```

```{r}
plotly_bacc_2017_longest_prds
```




## Kent Island 2018-2019

Let's visualize the switches/proportions for 2018-2019 files from Kent's Island (10 files total). Make sure you clear out your environment so you don't accidentally run this on Baccalieu-related objects or other sites you may have tried out.

```{r}
rm(list = ls(pattern = "^bacc"))
```

### Running plotting functions

Let's start by subsetting the metadata file so we can look at Kent's Island

```{r}

deployment_summary

```

Let's use the 10 files from 2018-2019.


**Proportions:**

```{r}
#subset the dataframe so we have less data to work with
kent_2018_props <- deg_props_2hrs %>%
  
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  
  select(bird_id, date, colony, start_time, end_time, prop_wet, deployment_period)

kent_2018_props
```


```{r}
plotly_kent2018_props <- plot_props(data = kent_2018_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (Kent Island 2018-2019)")

```


**Daily Proportions:**

```{r}
kent_2018_daily_props <- kent_2018_props %>%
  group_by(bird_id, date, colony, deployment_period) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")

plotly_kent2018_daily_props <- plot_daily_props(data = kent_2018_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (Kent Island 2018-2019)")

```


**Daily Switches:**

```{r}
#subset the dataframe so we have less data to work with
kent_2018_switches <- daily_switches %>%
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  select(bird_id, colony, date, daily_switches, deployment_period)

head(kent_2018_switches)
```

```{r}
plotly_kent2018_daily_switches <- plot_daily_switches(
  data = kent_2018_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for Kent Island 2018-2019")

```


**Longest Wet/Dry Periods**

```{r}
kent_2018_longest_prds <- deg_longest_prds %>%
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)
```

```{r}

plotly_kent_2018_longest_prds <- plot_longest_prds(
  data = kent_2018_longest_prds,
  plot_title = "Longest Wet and Dry Periods for Kent Island 2018-2019")

```


**Total Wet/Dry**

```{r}

kent_2018_total_wetdry <- deg_total_wetdry %>%
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plotly_kent_2018_total_times <- plot_total_times(kent_2018_total_wetdry, "Total Wet and Dry Times (Kent Island 2018-2019)")
```


### Final Plots 

```{r}
plotly_kent2018_props
```

```{r}
plotly_kent2018_daily_props
```

```{r}
plotly_kent2018_daily_switches
```

```{r}

plotly_kent_2018_total_times
```

```{r}
plotly_kent_2018_longest_prds
```


## Random subset

Now let's visualize the switches/proportions for 10 random files. Make sure you clear out your environment so you don't accidentally run this on Baccalieu-related or Kent-related objects or other sites you may have tried out.

```{r}
rm(list = ls(pattern = "^kent"), envir = .GlobalEnv)
rm(list = ls(pattern = "^bacc"), envir = .GlobalEnv)
rm(list = ls(pattern = "^random"), envir = .GlobalEnv)
```


### Running plotting functions

Let's start by subsetting the metadata file so we can look at 10 files from any random year and place.

```{r}
random_bird_ids <- metrics_md %>%
    
  distinct(bird_id) %>%
  sample_n(10) %>%
  
  pull(bird_id)

random_bird_ids

```


**Proportions:**

```{r}
#let's just put all the random subsets on the same year so we can visualize them better
# this means we have to determine the reference shift (let's make the smallest year = 1988)

random_props <- deg_props_2hrs %>%
    mutate(date = as.Date(date))

random_props <- deg_props_2hrs %>%
  
  filter(bird_id %in% random_bird_ids) %>%
  
  select(bird_id, colony, date, start_time, end_time, prop_wet, deployment_period) %>%
  
  group_by(bird_id) %>%
  
  mutate(
    min_year = min(year(date)),  # find the min date per bird_id
    relative_year = year(date) - min_year,
    fake_year = 1988 + relative_year,  # calculate the shifted year

    #fixing leap year issue (see note below)
    fake_date = make_date(
      year = fake_year,
      month = month(date),
      day = ifelse(month(date) == 2 & day(date) == 29 & !leap_year(fake_year), 28, day(date)))) %>%
  
  ungroup()


#make sure it filtered the same random files properly
print(unique(random_props$bird_id))
```

**NOTE:** Weird NA's kept popping up because of leap years I think? 2020 was a leap. 1989 does not have feb 29 so NA values were popping up

Before plotting let's make sure the dates all start at 1988 and that the dates are still relative for each bird_id

```{r}

random_props %>%
  group_by(bird_id) %>%
  summarize(
    min_fake_year = min(year(fake_date)),
    min_original_year = min(year(date))) %>%
  print()


```


```{r}
plotly_random_props <- plot_props_random(random_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (10 random files)")

```


**Daily Proportions:**

```{r}
random_daily_props <- random_props %>%
  group_by(bird_id, fake_date, colony, deployment_period) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")


plotly_random_daily_props <- plot_daily_props_random(data = random_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (10 random files)")

```


**Daily Switches:**

```{r}

#subset the dataframe so we have less data to work with

random_switches <- daily_switches %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id, date, daily_switches, colony, deployment_period) %>%
  
  group_by(bird_id) %>%  # ✅ Make sure each bird_id has its own reference year
  mutate(
    min_year = min(year(date), na.rm = TRUE),  
    relative_year = year(date) - min_year,  
    fake_year = 1988 + relative_year,
    fake_date = make_date(
      year = fake_year,
      month = month(date),
      day = ifelse(month(date) == 2 & day(date) == 29 & !leap_year(fake_year), 28, day(date))
    )
  ) %>%
  ungroup()



plotly_random_daily_switches <- plot_daily_switches_random(random_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for 10 Random Files")

```


**Longest Wet/Dry Periods**

```{r}
random_longest_prds <- deg_longest_prds %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)


plotly_random_longest_prds <- plot_longest_prds(
  data = random_longest_prds,
  plot_title = "Longest Wet and Dry Periods for 10 Random Files")

```


**Total Wet/Dry**

```{r}

random_total_wetdry <- deg_total_wetdry %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plotly_random_total_times <- plot_total_times(random_total_wetdry, "Total Wet and Dry Times for 10 Random Files")

```

### Final Plots 

```{r}
plotly_random_props
```

```{r}
plotly_random_daily_props
```

```{r}
plotly_random_daily_switches
```

```{r}
plotly_random_total_times
```

```{r}
plotly_random_longest_prds
```