# Visualizing Metrics
    
    From Katie's Notes:
    
    Within each 2hr period we need to assess:
    •	Proportion wet or time wet ((Count wets in 2hr period * 30sec) / 72000sec)
    •	Number of switches between 0 and non-zero value
    
    •	Total time when entire 10min sampling interval was dry (count 0s * 10min)
    •	Total time when entire 10min sampling interval was wet (count >=1 * 10min)
    
    •	Longest continuous time dry (longest number of consecutive rows with wets==0  * 10min)
    •	Longest continuous time wet (longest number of consecutive rows with wets>=1 * 10min)


I start with setting up some plotting functions so we can look at different subsets of our data easily. Then I apply those plotting functions to various sets of the DEG files (6 files) to visualize the metrics we discussed (for Baccalieu, Kent's Island, and random files).

```{r include=FALSE}
#Load the plotting functions 
source(here::here("utils/plotting_functions.R"), echo=TRUE)
```
_____________________________________________________________________


Let's see what we have to work with per colony/deployment_periods
```{r}
deployment_summary
```



## Baccalieu Island 2017-2018

Let's start by visualizing switches/proportions for 2017-2018 files from Baccalieu (6 files total)

### Running plotting functions

**Proportions:**

```{r}
#subset the dataframe so we have less data to work with
bacc_2017_props <- deg_props_2hrs %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id, date, start_time, end_time, prop_wet)

head(bacc_2017_props)
```


```{r}
plot_bacc2017_props <- plot_props(data = bacc_2017_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (Baccalieu Island 2017-2018)")

```


This kind of hurts to look at so we can also try aggregating the data to days taking the sum of prop_wet.

**Daily Proportions:**

```{r}
bacc_2017_daily_props <- bacc_2017_props %>%
  group_by(bird_id, date) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")


plot_bacc2017_daily_props <- plot_daily_props(data = bacc_2017_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (Baccalieu Island 2017-2018)")

```


**Daily Switches:**

```{r}
#subset the dataframe so we have less data to work with
bacc_2017_switches <- daily_switches %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id, date, daily_switches)

head(bacc_2017_switches)
```

```{r}
plot_bacc2017_daily_switches <- plot_daily_switches(
  data = bacc_2017_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for Baccalieu Island (2017-2018)")

```


**Longest Wet/Dry Periods**

```{r}
bacc_2017_longest_prds <- deg_longest_prds %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)
```

```{r}
plot_bacc_2017_longest_prds <- plot_longest_prds(
  data = bacc_2017_longest_prds,
  plot_title = "Longest Wet and Dry Periods - Baccalieu Island 2017-2018")

```



**Total Wet/Dry**

```{r}

bacc_2017_total_wetdry <- deg_total_wetdry %>%
  filter(colony == "Baccalieu - Ned Walsh",
         deployment_period == "2017-2018") %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plot_bacc_2017_total_times <- plot_total_times(bacc_2017_total_wetdry, "Total Wet and Dry Times (Baccalieu Island, 2017-2018)")
```


### Final Plots 

```{r}
plot_bacc2017_props
```

```{r}
plot_bacc2017_daily_props
```

```{r}
plot_bacc2017_daily_switches
```

```{r}
plot_bacc_2017_total_times
```

```{r}
plot_bacc_2017_longest_prds
```




## Kent Island 2018-2019

Let's visualize the switches/proportions for 2018-2019 files from Kent's Island (10 files total). Make sure you clear out your environment so you don't accidentally run this on Baccalieu-related objects or other sites you may have tried out.

```{r}
rm(list = ls(pattern = "^bacc"))
```

### Running plotting functions

Let's start by subsetting the metadata file so we can look at Kent's Island

```{r}

deployment_summary

```

Let's use the 10 files from 2018-2019.


**Proportions:**

```{r}
#subset the dataframe so we have less data to work with
kent_2018_props <- deg_props_2hrs %>%
  
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  
  select(bird_id, date, start_time, end_time, prop_wet)

kent_2018_props
```


```{r}
plot_kent2018_props <- plot_props(data = kent_2018_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (Kent Island 2018-2019)")

```


**Daily Proportions:**

```{r}
kent_2018_daily_props <- kent_2018_props %>%
  group_by(bird_id, date) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")

plot_kent2018_daily_props <- plot_daily_props(data = kent_2018_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (Kent Island 2018-2019)")

```


**Daily Switches:**

```{r}
#subset the dataframe so we have less data to work with
kent_2018_switches <- daily_switches %>%
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  select(bird_id, date, daily_switches)

head(kent_2018_switches)
```

```{r}
plot_kent2018_daily_switches <- plot_daily_switches(
  data = kent_2018_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for Kent Island 2018-2019")

```


**Longest Wet/Dry Periods**

```{r}
kent_2018_longest_prds <- deg_longest_prds %>%
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)
```

```{r}

plot_kent_2018_longest_prds <- plot_longest_prds(
  data = kent_2018_longest_prds,
  plot_title = "Longest Wet and Dry Periods for Kent Island 2018-2019")

```


**Total Wet/Dry**

```{r}

kent_2018_total_wetdry <- deg_total_wetdry %>%
  filter(colony == "Kent",
         deployment_period == "2018-2019") %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plot_kent_2018_total_times <- plot_total_times(kent_2018_total_wetdry, "Total Wet and Dry Times (Kent Island 2018-2019)")
```


### Final Plots 

```{r}
plot_kent2018_props
```

```{r}
plot_kent2018_daily_props
```

```{r}
plot_kent2018_daily_switches
```

```{r}

plot_kent_2018_total_times
```

```{r}
plot_kent_2018_longest_prds
```


## Random subset

Now let's visualize the switches/proportions for 10 random files
Make sure you clear out your environment so you don't accidentally run this on Baccalieu- or Kent-related objects or other sites you may have tried out.

```{r}
rm(list = ls(pattern = "^kent"), envir = .GlobalEnv)
rm(list = ls(pattern = "^bacc"), envir = .GlobalEnv)
```


### Running plotting functions

Let's start by subsetting the metadata file so we can look at 10 files from any random year and place.

```{r}
random_bird_ids <- metrics_md %>%
    
  distinct(bird_id) %>%
  sample_n(10) %>%
  
  pull(bird_id)

random_bird_ids

```
**Proportions:**

```{r}

#subset the dataframe so we have less data to work with
random_props <- deg_props_2hrs %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id, colony, date, start_time, end_time, prop_wet)

#make sure it filtered the same random files properly
print(unique(random_props$bird_id))

```


```{r}
plot_random_props <- plot_props(data = random_props, 
                                  y_var = "prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Proportion of Wet Time Across Deployments Per 2hr bins) - (10 random files)")

```


**Daily Proportions:**

```{r}
random_daily_props <- random_props %>%
  group_by(bird_id, date) %>%
  summarize(daily_prop_wet = mean(prop_wet, na.rm = TRUE), .groups = "drop")

plot_random_daily_props <- plot_daily_props(data = random_daily_props, 
                                  y_var = "daily_prop_wet", 
                                  y_label = "Proportion", 
                                  plot_title = "Mean Proportion of Wet Periods (2hr bins) Per Day (10 random files)")

```


**Daily Switches:**

```{r}
#subset the dataframe so we have less data to work with
random_switches <- daily_switches %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id, date, daily_switches)

#make sure it filtered the same random files
print(unique(random_switches$bird_id))
```

```{r}
plot_random_daily_switches <- plot_daily_switches(
  data = random_switches,
  y_var = "daily_switches",
  y_label = "Sum of Switches",
  plot_title = "Daily Sum of Switches between Wet (wet>0) and Dry (wet=0) States for 10 Random Files")

```


**Longest Wet/Dry Periods**

```{r}
random_longest_prds <- deg_longest_prds %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id,
         longest_dry_time, 
         longest_dry_start, 
         longest_dry_end, 
         longest_wet_time,
         longest_wet_start, 
         longest_wet_end)

```

```{r}

plot_random_longest_prds <- plot_longest_prds(
  data = random_longest_prds,
  plot_title = "Longest Wet and Dry Periods for 10 Random Files")

```


**Total Wet/Dry**

```{r}

random_total_wetdry <- deg_total_wetdry %>%
  filter(bird_id %in% random_bird_ids) %>%
  select(bird_id,
         total_time_wet,
         total_time_dry)


plot_random_total_times <- plot_total_times(random_total_wetdry, "Total Wet and Dry Times for 10 Random Files")

```


### Final Plots 

```{r}
plot_random_props
```

```{r}
plot_random_daily_props
```

```{r}
plot_random_daily_switches
```

```{r}
plot_random_total_times
```

```{r}
plot_random_longest_prds
```