# Hidden Markov Modelling (HMM)


Datasets of interest are `deg_props_2hrs` and `deg_switches`
Other datasets we migth use are `deg_total_dry` and `deg_longest_prds_time`

## Data Prep

```{r include=FALSE}

rm(list = setdiff(ls(), c("deg_props_2hrs", "deg_switches", "deg_longest_prds", "deg_total_dry")))

source(here::here("project_settings.R"), echo=FALSE)

```


Now let's clip the files, we don't want files that have the fully dry days when the bird is in the burrow in the non-breeding analysis. 


### Clip Deployment Day

First we want to clip each bird's data to start logging dates the day after the tracking device was deployed. Let's clip the first date listed for every bird ID, unless the bird ID starts logging before August 15, then we should clip every date up until August 15. 

Here is a function to clip the first day from both datasets
```{r}

clip_first_days <- function(data) {
  
  first_days <- data %>%  #find first day, 
    group_by(bird_id) %>%
    summarize(first_date = min(date, na.rm = TRUE), .groups = "drop") %>%
    
    # if first day is before Aug 15 then use Aug 15 as first day
    mutate(clip_date = if_else(first_date < as.Date(paste0(year(first_date), "-08-15")),
                               as.Date(paste0(year(first_date), "-08-15")),
                               first_date + 1))

    data_filtered <- data %>%  #filter based on first day
    left_join(first_days, by = "bird_id") %>%
    filter( date >= clip_date) %>%
    select(-first_date, -clip_date)
}

```


Run the function in the proportions dataset 
```{r}
deg_props_2hrs_nb <- clip_first_days(deg_props_2hrs)

head(deg_props_2hrs_nb)
```

and then switches dataset:
```{r}

deg_switches_nb <- clip_first_days(deg_switches)

head(deg_switches_nb)

```

Then a function to get the first dates before and after clipping
```{r}
get_clip_summary <- function(data_before, data_after) {
  summary <- data_after %>%
    group_by(bird_id) %>%
    summarize(first_date_after = min(date, na.rm = TRUE)) %>%
    left_join(
      data_before %>%
        group_by(bird_id) %>%
        summarize(first_date_before = min(date, na.rm = TRUE)),
      by = "bird_id"
    ) 
  
  return(summary)
}

```

```{r}
# For deg_props
clip_summary_props <- get_clip_summary(deg_props_2hrs, deg_props_2hrs_nb)
print(clip_summary_props)
```

```{r}
# For deg_switches
clip_summary_switches <- get_clip_summary(deg_switches, deg_switches_nb)
print(clip_summary_switches)
```

All looks good for 274 files.


### Clip Burrow-Use

Now on the retrieval end of the datasets we want to stop the day before the devices were retrieved. We can use the dataframe `LHSP_burrowUse_returnOnly.rds` for this. We're looking for the ID_yr (corresponding to bird_id) and datetime_start columns to clip the end of our datasets and exclude fully dry days from our analysis when the device was retrieved.

```{r}

# read in the burrow use data
burrow_use <- readRDS("C:/Users/BoschJ/Desktop/wet-dry_analysis/data/LHSP_burrowUse_returnOnly.rds")

burrow_use <- burrow_use %>%
  
  mutate(  
    
    #isolate bird ID into a column called bird_id
    bird_id = str_extract(ID_yr, "^[^_]+"),
    
    #split the datetime columns up to isolate dates
    start_date = as.Date(datetime_start)) %>%
  
  group_by(bird_id) %>%
  summarize(start_date = min(start_date, na.rm = TRUE)) %>%
  ungroup() %>%
    
  select(  #keep only ID_year, datetime_start, datetime_end columns
    bird_id, start_date)

print(burrow_use)

```


Let's build a function to clip burrow use from any dataframe we would like:

```{r}

clip_burrow_use <- function(data, burrow_use) {
  
  #merge summarized burrow-use dates
  data_joined <- data %>%
    left_join(burrow_use, by = "bird_id")
  
  
  #filter out burrow-use periods
  data_filtered <- data_joined %>%
    filter(is.na(start_date) | date <= start_date) %>%
    select(-start_date)    #remove columns 
  
  return(data_filtered)
}

```


```{r}

#clip burrow-use periods from the proportions df, named 'nb' for non-breeding
deg_props_2hrs_nb_clipped <- clip_burrow_use(deg_props_2hrs, burrow_use)


print(deg_props_2hrs_nb_clipped)

```


```{r}
#clip burrow-use periods from the proportions df, named 'nb' for non-breeding
deg_switches_nb_clipped <- clip_burrow_use(deg_switches, burrow_use)

print(deg_switches_nb_clipped)

```

Now a function to check the last date for each bird in the clipped data

```{r}
check_last_day <- function(original_data, clipped_data, burrow_use) {
  
  # get the earliest burrow start date for each bird
  burrow_use_summary <- burrow_use %>%
    group_by(bird_id) %>%
    summarize(start_date = min(start_date, na.rm = TRUE)) %>%
    ungroup()
  
  # get the last date before clipping
  last_day_before_clip <- original_data %>%
    group_by(bird_id) %>%
    summarize(last_day_before_clip = max(date, na.rm = TRUE)) %>%
    ungroup()
  
  # get the last date after clipping
  last_day_after_clip <- clipped_data %>%
    group_by(bird_id) %>%
    summarize(last_day_after_clip = max(date, na.rm = TRUE)) %>%
    ungroup()
  
  # then just merge all three 
  summary_table <- last_day_before_clip %>%
    left_join(burrow_use_summary, by = "bird_id") %>%
    left_join(last_day_after_clip, by = "bird_id") %>%
    mutate(is_clipped_correctly = last_day_after_clip == start_date)
  
  return(summary_table)
}

```

```{r}
# summary for the proportions dataset
comparison_props <- check_last_day(deg_props_2hrs_nb, deg_props_2hrs_nb_clipped, burrow_use)
print(comparison_props)
```

```{r}
# summary for the switches dataset
comparison_switches <- check_last_day(deg_switches_nb, deg_switches_nb_clipped, burrow_use)
print(comparison_switches)
```

Now we need to consider that there may be a few birds that never spent a full day in the burrow before they were retrieved, so you those datasets should be stopped the day before they were retrieved. See there are NAs in the summary.



1)  create a single clean data frame containing data for every file


2)  Run a 2state HMMM, have to run the following for the HMM:

    prepData()
    
    define variables:
    
    stateNames <- dry, wet
    stepPar <- (inital mean for states based on prop_wet)
    anglePar <- (no angle parameter for our analysis)
    
    fitHMM()
    plotSates() - visualize the model


3) test with different binning intervals (2hr, 5hr?)

4) test covariates

  * wetness level (0-20) as a covariate.
  * maybe seasonality or a day/night indicator?

REFERENCES: 
https://cran.r-project.org/web/packages/momentuHMM/vignettes/momentuHMM.pdf

https://mjones029.github.io/Tutorials/HMM_tutorial.html