# Hidden Markov Modelling (HMM)


Datasets of interest are `deg_props_2hrs` and `deg_switches`
Other datasets we migth use are `deg_total_dry` and `deg_longest_prds_time`

## Data Prep

```{r include=FALSE}
source(here::here("project_settings.R"), echo=TRUE)
```

### Clip Burrow-Use

Now let's clip the files for the burrow use periods, we don't want files that have the fully dry days when the bird is in the burrow in the non-breeding analysis. We're looking for the ID_yr, datetime_start, datetime_end columns to clip the end of our datasets and exclude fully dry days from our analysis.

```{r}

# read in the burrow use data
burrow_use <- readRDS("C:/Users/BoschJ/Desktop/wet-dry_analysis/data/LHSP_burrowUse_returnOnly.rds")

burrow_use <- burrow_use %>%
  
  mutate(  
    
    #isolate bird ID into a column called bird_id
    bird_id = str_extract(ID_yr, "^[^_]+"),
    
    #split the datetime columns up to isolate dates
    start_date = as.Date(datetime_start),
    end_date = as.Date(datetime_end)) %>%
  
  select(  #keep only ID_year, datetime_start, datetime_end columns
    bird_id, start_date, end_date) 

head(burrow_use, 100)

```


Let's build a function to clip burrow use from any dataframe we would like:

```{r}

clip_burrow_use <- function(data, burrow_use) {
  
  #summarize earliest and latest burrow use
  burrow_use_summary <- burrow_use %>%
    group_by(bird_id) %>%
    summarize(
      start_date = min(start_date, na.rm = TRUE),
      end_date = max(end_date, na.rm = TRUE))
  
  #merge summarized burrow-use dates
  data_joined <- data %>%
    left_join(burrow_use_summary, by = "bird_id")
  
  
  #filter out burrow-use periods
  data_filtered <- data_joined %>%
    filter(is.na(start_date) | date < start_date | date > end_date) %>%
    select(-start_date, -end_date)    #remove columns 
  
  return(data_filtered)
}

```

And here is a final df with all the summarized burrow use periods
```{r}

 burrow_use_summary<- burrow_use %>%
    group_by(bird_id) %>%
    summarize(
      start_date = min(start_date, na.rm = TRUE),
      end_date = max(end_date, na.rm = TRUE))

burrow_use_summary
```

If bird_id in `burrow_use` matches bird_id in the df's below, then clip the df's to exclude any dates that fall in between `start_date` and `end_date` of `burrow_use`. 
```{r}

#clip burrow-use periods from the proportions df, named 'nb' for non-breeding
deg_props_2hrs_nb <- clip_burrow_use(deg_props_2hrs, burrow_use)


print(deg_props_2hrs_nb)

```

Check to make sure it removed burrow-use periods from all 274 files and make sure rows were actually removed.

```{r}
summary_counts <- deg_props_2hrs %>%
  count(bird_id, name = "before") %>%
  left_join(deg_props_2hrs_nb %>% count(bird_id, name = "after"), by = "bird_id") %>%
  mutate(removed = before - replace_na(after, 0))

print(summary_counts)
```

```{r}

#clip burrow-use periods from the proportions df, named 'nb' for non-breeding
deg_switches_nb <- clip_burrow_use(deg_switches, burrow_use)

print(deg_switches_nb)

```

```{r}

# do the check for switches
summary_counts <- deg_switches %>%
  count(bird_id, name = "before") %>%
  left_join(deg_switches_nb %>% count(bird_id, name = "after"), by = "bird_id") %>%
  mutate(removed = before - replace_na(after, 0))

print(summary_counts)
```


### Clip Deployment Day

We can also clip the first day the device was deployed because we trap the birds in the burrows so we can catch them, we know for sure they are dry this day.

Here is a function to clip the first day from both datasets
```{r}

clip_first_days <- function(data) {
  
  first_days <- data %>%
    group_by(bird_id) %>%
    summarize(first_date = min(date, na.rm = TRUE), .groups = "drop")
  
  data_filtered <- data %>%
    left_join(first_days, by = "bird_id") %>%
    filter( date > first_date) %>%
    select(-first_date)
}

```

Run the function in the proportions dataset 

```{r}
clip_first_day_summary <- deg_props_2hrs_nb %>%
  group_by(bird_id) %>%
  summarize(first_date_before = min(date, na.rm = TRUE))
```


```{r}
deg_props_2hrs_nb <- clip_first_days(deg_props_2hrs_nb)

head(deg_props_2hrs_nb)
```
```{r}
clip_first_day_summary <- clip_first_day_summary %>%
  left_join(
    deg_props_2hrs_nb %>%
      group_by(bird_id) %>%
      summarize(first_date_after = min(date, na.rm = TRUE)),
    by = "bird_id") %>%
  print()
```


```{r}

deg_switches_nb <- clip_first_day(deg_switches_nb)

head(deg_switches_nb)

```


Run the function on the switches dataset:
```{r}
clip_first_day_summary <- deg_switches_nb %>%
  group_by(bird_id) %>%
  summarize(first_date_before = min(date, na.rm = TRUE))
```


```{r}
deg_switches_nb <- clip_first_days(deg_switches_nb)

head(deg_switches_nb)
```

```{r}
clip_first_day_summary <- clip_first_day_summary %>%
  left_join(
    deg_switches_nb %>%
      group_by(bird_id) %>%
      summarize(first_date_after = min(date, na.rm = TRUE)),
    by = "bird_id") %>%
  print()
```
274 files processed for each dataset, should be good then. 
In the future can make a mor erobust way of checking, but I think this is good for now.


1)  create a single clean data frame containing data for every file


2)  Run a 2state HMMM, have to run the following for the HMM:

    prepData()
    
    define variables:
    
    stateNames <- dry, wet
    stepPar <- (inital mean for states based on prop_wet)
    anglePar <- (no angle parameter for our analysis)
    
    fitHMM()
    plotSates() - visualize the model


3) test with different binning intervals (2hr, 5hr?)

4) test covariates

  * wetness level (0-20) as a covariate.
  * maybe seasonality or a day/night indicator?

REFERENCES: 
https://cran.r-project.org/web/packages/momentuHMM/vignettes/momentuHMM.pdf

https://mjones029.github.io/Tutorials/HMM_tutorial.html